---
title: "plinking_duck LD benchmark: upstream vs sounkou fork"
output:
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
if (dir.exists(".sync/plinking_duck")) {
  repo_root <- normalizePath(".")
} else {
  repo_root <- normalizePath("..")
}
knitr::opts_knit$set(root.dir = repo_root)
setwd(repo_root)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(width = 120)
```

## Purpose

This benchmark compares local builds of:

- upstream `teaguesterling/plinking_duck` cloned at `.sync/plinking_duck`
- `sounkou-bioinfo/plinking_duck` fork cloned at `.sync/plinking_duck_sounkou`

The fork adds signed `R`, projection-aware LD output, and a bounded cached
popcount path for no-missing regional LD scans.  All timings below are measured
with `/usr/bin/time -v`, so the table includes wall time and peak resident set
size.

## Machine, inputs, and build artifacts

```{r config}
threads <- as.integer(Sys.getenv("PLINKING_BENCH_THREADS", "19"))
n_reps <- as.integer(Sys.getenv("PLINKING_BENCH_REPS", "3"))

paths <- list(
  upstream_dir = normalizePath(".sync/plinking_duck"),
  fork_dir = normalizePath(".sync/plinking_duck_sounkou"),
  upstream_duckdb = normalizePath(".sync/plinking_duck/build/release/duckdb"),
  fork_duckdb = normalizePath(".sync/plinking_duck_sounkou/build/release/duckdb"),
  pgen = normalizePath(".sync/plink_resources/1kg_chr22_hg38/chr22_hg38.pgen"),
  pvar = normalizePath(".sync/plink_resources/1kg_chr22_hg38/chr22_hg38.pvar"),
  psam = normalizePath(".sync/plink_resources/1kg_chr22_hg38/chr22_hg38.psam"),
  plink2 = normalizePath(".sync/plink2/plink2", mustWork = FALSE)
)

stopifnot(file.exists(paths$upstream_duckdb), file.exists(paths$fork_duckdb),
          file.exists(paths$pgen), file.exists(paths$pvar), file.exists(paths$psam))

cmd_out <- function(cmd, args = character()) {
  paste(system2(cmd, args, stdout = TRUE, stderr = TRUE), collapse = "\n")
}

git_head <- function(dir) cmd_out("git", c("-C", dir, "rev-parse", "--short", "HEAD"))
file_mb <- function(path) round(file.info(path)$size / 1024^2, 2)

info <- data.frame(
  item = c("detected CPUs", "benchmark threads", "repetitions", "upstream HEAD", "fork HEAD",
           "upstream duckdb MB", "fork duckdb MB", "PGEN MB", "region", "window kb", "r2 threshold"),
  value = c(cmd_out("nproc"), threads, n_reps, git_head(paths$upstream_dir), git_head(paths$fork_dir),
            file_mb(paths$upstream_duckdb), file_mb(paths$fork_duckdb), file_mb(paths$pgen),
            "22:11300000-11400000", 100, 0.2)
)
knitr::kable(info)
```

## pgenlibr header/linkage audit

This answers the linkage question directly: the installed R package exposes R
functions and a shared object, but it does **not** install public C/C++ headers
under `include/`. The CRAN source tarball contains PLINK2/pgenlib headers under
`tools/include`, but those are source-build internals rather than a stable
`LinkingTo: pgenlibr` API. For a DuckDB extension, linking directly to vendored
PLINK2 `pgenlib` is the right boundary.

```{r pgenlibr-audit}
pgenlibr_installed <- requireNamespace("pgenlibr", quietly = TRUE)
pgenlibr_pkg_path <- if (pgenlibr_installed) system.file(package = "pgenlibr") else NA_character_
pgenlibr_installed_include <- if (pgenlibr_installed) file.path(pgenlibr_pkg_path, "include") else NA_character_
pgenlibr_source_headers <- list.files(".sync/pgenlibr", pattern = "\\.(h|hpp)$", recursive = TRUE, full.names = FALSE)
pgenlib_source_headers <- grep("pgenlib|plink2|pvar", pgenlibr_source_headers, value = TRUE)

audit <- data.frame(
  check = c("pgenlibr installed", "installed package path", "installed include/ exists",
            "source headers matching pgenlib/plink2/pvar"),
  value = c(pgenlibr_installed, pgenlibr_pkg_path, dir.exists(pgenlibr_installed_include),
            length(pgenlib_source_headers))
)
knitr::kable(audit)
head(pgenlib_source_headers, 20)
```

## Benchmark helpers

```{r helpers}
parse_elapsed <- function(x) {
  if (length(x) == 0 || is.na(x)) return(NA_real_)
  parts <- as.numeric(strsplit(x, ":", fixed = TRUE)[[1]])
  if (length(parts) == 3) return(parts[1] * 3600 + parts[2] * 60 + parts[3])
  if (length(parts) == 2) return(parts[1] * 60 + parts[2])
  parts[1]
}

field <- function(lines, prefix) {
  hit <- grep(paste0("^\\s*", prefix), lines, value = TRUE)
  if (!length(hit)) return(NA_character_)
  sub(paste0("^\\s*", prefix, ":\\s*"), "", hit[1])
}

run_timed <- function(label, exe, args, rep = NA_integer_) {
  stdout_file <- tempfile("bench_stdout_")
  stderr_file <- tempfile("bench_stderr_")
  # system2() still invokes a shell on Unix, so quote every executable/argument;
  # SQL strings contain semicolons and parentheses.
  status <- system2("/usr/bin/time", c("-v", shQuote(exe), shQuote(args)), stdout = stdout_file, stderr = stderr_file)
  if (is.null(status)) status <- 0L
  out <- if (file.exists(stdout_file)) readLines(stdout_file, warn = FALSE) else character()
  err <- if (file.exists(stderr_file)) readLines(stderr_file, warn = FALSE) else character()

  elapsed_raw <- field(err, "Elapsed \\(wall clock\\) time \\(h:mm:ss or m:ss\\)")
  data.frame(
    label = label,
    rep = rep,
    status = as.integer(status),
    elapsed_sec = parse_elapsed(elapsed_raw),
    user_sec = as.numeric(field(err, "User time \\(seconds\\)")),
    sys_sec = as.numeric(field(err, "System time \\(seconds\\)")),
    cpu_percent = as.numeric(sub("%", "", field(err, "Percent of CPU this job got"), fixed = TRUE)),
    max_rss_mb = as.numeric(field(err, "Maximum resident set size \\(kbytes\\)")) / 1024,
    stdout = paste(out, collapse = "\\n"),
    stderr_tail = paste(unique(c(grep("Error|Exception|Binder", err, value = TRUE), tail(err, 12))), collapse = "\\n"),
    stringsAsFactors = FALSE
  )
}

ld_from <- function(extra = "") {
  sprintf(
    "FROM plink_ld('%s', pvar := '%s', psam := '%s', region := '22:11300000-11400000', window_kb := 100, r2_threshold := 0.2%s)",
    paths$pgen, paths$pvar, paths$psam, extra
  )
}

duck_sql <- function(select, extra = "") {
  paste0("SET threads=", threads, "; SET plinking_max_threads=", threads, "; ", select, " ", ld_from(extra), ";")
}

run_duckdb <- function(repo, exe, scenario, select, extra = "", reps = n_reps) {
  rows <- vector("list", reps)
  for (i in seq_len(reps)) {
    sql <- duck_sql(select, extra)
    rows[[i]] <- run_timed(paste(repo, scenario, sep = ":"), exe, c("-csv", "-c", sql), rep = i)
    rows[[i]]$repo <- repo
    rows[[i]]$scenario <- scenario
  }
  do.call(rbind, rows)
}

summarise_bench <- function(x) {
  ok <- x[x$status == 0, ]
  parts <- split(ok, paste(ok$scenario, ok$repo, sep = "\r"))
  out <- lapply(parts, function(d) {
    data.frame(
      scenario = d$scenario[1],
      repo = d$repo[1],
      n = nrow(d),
      median_elapsed_sec = median(d$elapsed_sec),
      min_elapsed_sec = min(d$elapsed_sec),
      median_max_rss_mb = median(d$max_rss_mb),
      median_cpu_percent = median(d$cpu_percent),
      output = gsub("\\n", "; ", d$stdout[1], fixed = TRUE),
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, out)
}
```

## Run DuckDB LD benchmarks

Scenarios:

- `count_pairs`: compute all windowed LD pairs and return only the count.
- `sum_r2`: compute and aggregate `R2` only.
- `sum_r2_dprime`: compute `R2` plus `D_PRIME`, validating that the optimized
  path preserves the legacy statistic when requested.

```{r run-duckdb-bench}
benchmarks <- list(
  run_duckdb("upstream", paths$upstream_duckdb, "count_pairs", "SELECT COUNT(*)"),
  run_duckdb("fork", paths$fork_duckdb, "count_pairs", "SELECT COUNT(*)"),
  run_duckdb("upstream", paths$upstream_duckdb, "sum_r2", "SELECT SUM(R2)"),
  run_duckdb("fork", paths$fork_duckdb, "sum_r2", "SELECT SUM(R2)"),
  run_duckdb("upstream", paths$upstream_duckdb, "sum_r2_dprime", "SELECT COUNT(*), SUM(R2), SUM(D_PRIME)"),
  run_duckdb("fork", paths$fork_duckdb, "sum_r2_dprime", "SELECT COUNT(*), SUM(R2), SUM(D_PRIME)")
)
raw <- do.call(rbind, benchmarks)
raw_out <- file.path("benchmarks", "plinking_duck_ld_benchmark_raw.csv")
write.csv(raw, raw_out, row.names = FALSE)
knitr::kable(raw[, c("scenario", "repo", "rep", "status", "elapsed_sec", "user_sec", "sys_sec", "cpu_percent", "max_rss_mb")], digits = 3)
```

## Summary and speedup

```{r summary}
summary <- summarise_bench(raw)
summary_out <- file.path("benchmarks", "plinking_duck_ld_benchmark_summary.csv")
write.csv(summary, summary_out, row.names = FALSE)
knitr::kable(summary[, c("scenario", "repo", "n", "median_elapsed_sec", "min_elapsed_sec", "median_max_rss_mb", "median_cpu_percent")], digits = 3)

up <- summary[summary$repo == "upstream", c("scenario", "median_elapsed_sec", "median_max_rss_mb")]
fk <- summary[summary$repo == "fork", c("scenario", "median_elapsed_sec", "median_max_rss_mb")]
names(up) <- c("scenario", "upstream_elapsed", "upstream_rss")
names(fk) <- c("scenario", "fork_elapsed", "fork_rss")
compare <- merge(up, fk, by = "scenario")
compare$speedup_x <- compare$upstream_elapsed / compare$fork_elapsed
compare$rss_delta_mb <- compare$fork_rss - compare$upstream_rss
knitr::kable(compare, digits = 3)
```

## Full all_hg38 1000G benchmark (real large resource)

This run uses the real full-resource files at repository root:

- `./all_hg38.pgen.zst` as the source PGEN (decompressed to `./all_hg38.pgen` because pgenlib rejects `.pgen.zst` magic bytes)
- `./all_hg38_rs.pvar.zst` as the source PVAR, with `./all_hg38_rs.slim.pvar` derived by retaining the real `#CHROM POS ID REF ALT` columns and dropping the huge INFO field
- `./hg38_corrected.psam` as sample metadata

The benchmark interval is `22:11300000-16300000` (~5 Mb; 91,882 variants),
with a 100 kb LD window and `r2_threshold = 0.2`. This is intentionally much
larger than the earlier 100 kb smoke benchmark and exercises the full all-hg38
PGEN/PVAR metadata scale. The CSV below is produced from the `/usr/bin/time -v`
run; re-running it takes several minutes and requires the decompressed 8.9 GB
PGEN plus the 2.0 GB slim PVAR.

Commands used to prepare real inputs (not committed because they are large):

```sh
zstd -T0 -d -k all_hg38.pgen.zst -o all_hg38.pgen
zstd -dc -T0 all_hg38_rs.pvar.zst | \
  awk 'BEGIN{FS=OFS="\t"} /^##/ {next} /^#CHROM/ {print "#CHROM","POS","ID","REF","ALT"; next} !/^#/ {print $1,$2,$3,$4,$5}' \
  > all_hg38_rs.slim.pvar
```

```{r all-hg38-bench-results}
all_hg38 <- read.csv(file.path("benchmarks", "plinking_duck_ld_benchmark_all_hg38.csv"))
knitr::kable(all_hg38, digits = 3)

all_hg38_compare <- merge(
  all_hg38[all_hg38$repo == "upstream", c("scenario", "elapsed_sec", "max_rss_mb")],
  all_hg38[all_hg38$repo == "fork", c("scenario", "elapsed_sec", "max_rss_mb")],
  by = "scenario", suffixes = c("_upstream", "_fork")
)
all_hg38_compare$speedup_x <- all_hg38_compare$elapsed_sec_upstream / all_hg38_compare$elapsed_sec_fork
all_hg38_compare$rss_delta_mb <- all_hg38_compare$max_rss_mb_fork - all_hg38_compare$max_rss_mb_upstream
knitr::kable(all_hg38_compare, digits = 3)
```

Note: the fork now uses native PVAR region pushdown for `plink_ld()` when a `region` is supplied and shares immutable pgenlib metadata across workers. On the real all-hg38 resource this cuts peak RSS substantially; elapsed time is now dominated by the LD kernel / PGEN reads rather than all-PVAR or per-worker pgen metadata materialization.

## Output agreement

```{r output-agreement}
outputs <- summary[, c("scenario", "repo", "output")]
knitr::kable(outputs)
```

## Signed R smoke test

The fork should expose signed `R`; upstream should fail because the column does
not exist.

```{r signed-r}
signed_sql <- duck_sql("SELECT R, R2", "")
signed_sql <- sub(";$", " LIMIT 5;", signed_sql)
signed_upstream <- run_timed("upstream:signed_R", paths$upstream_duckdb, c("-csv", "-c", signed_sql), rep = 1)
signed_fork <- run_timed("fork:signed_R", paths$fork_duckdb, c("-csv", "-c", signed_sql), rep = 1)
signed <- rbind(
  transform(signed_upstream, repo = "upstream"),
  transform(signed_fork, repo = "fork")
)
knitr::kable(signed[, c("repo", "status", "elapsed_sec", "max_rss_mb", "stdout", "stderr_tail")], digits = 3)
```

## Population-weighted LD smoke benchmark

This exercises the GAUSS-like population-weighted path added to the fork. The
`.psam` file has a `SuperPop` column; each superpopulation below contributes 20%
of the total sample weight regardless of how many samples it has in the panel.

```{r population-weighted}
weighted_sql <- duck_sql(
  "SELECT COUNT(*) AS n_pairs, SUM(R2) AS sum_r2, SUM(R) AS sum_r",
  ", population_column := 'SuperPop', population_weights := 'AFR=0.20,AMR=0.20,EAS=0.20,EUR=0.20,SAS=0.20'"
)
weighted_fork <- run_timed("fork:population_weighted", paths$fork_duckdb, c("-csv", "-c", weighted_sql), rep = 1)
weighted_display <- weighted_fork
weighted_display$stdout <- gsub("\\n", "; ", weighted_display$stdout, fixed = TRUE)
knitr::kable(weighted_display[, c("label", "status", "elapsed_sec", "user_sec", "sys_sec", "cpu_percent", "max_rss_mb", "stdout")], digits = 3)
```

## PLINK2 CLI reference point

This is not a DuckDB implementation, but it is a useful external reference for
how fast PLINK2's native LD kernel is on the same input and region.

```{r plink2-reference}
run_plink2 <- function(reps = n_reps) {
  if (!file.exists(paths$plink2)) return(data.frame())
  prefix <- sub("\\.pgen$", "", paths$pgen)
  rows <- vector("list", reps)
  for (i in seq_len(reps)) {
    out_prefix <- tempfile("plink2_ld_")
    args <- c(
      "--pfile", prefix,
      "--chr", "22", "--from-bp", "11300000", "--to-bp", "11400000",
      "--ld-window-kb", "100", "--ld-window-r2", "0.2",
      "--threads", as.character(threads),
      "--r-unphased", "ref-based", "cols=id,ref,alt",
      "--out", out_prefix
    )
    rows[[i]] <- run_timed("plink2:ld", paths$plink2, args, rep = i)
    rows[[i]]$repo <- "plink2"
    rows[[i]]$scenario <- "plink2_r_unphased"
  }
  do.call(rbind, rows)
}
plink2_raw <- run_plink2()
if (nrow(plink2_raw)) {
  write.csv(plink2_raw, file.path("benchmarks", "plinking_duck_ld_benchmark_plink2.csv"), row.names = FALSE)
  knitr::kable(plink2_raw[, c("scenario", "repo", "rep", "status", "elapsed_sec", "user_sec", "sys_sec", "cpu_percent", "max_rss_mb")], digits = 3)
} else {
  cat("PLINK2 binary not found; skipped.\n")
}
```

## Notes

- The fork trades memory for speed on bounded regional scans by caching packed
  genotypes per worker. The benchmark's RSS delta is therefore expected.
- The cached fast path is only used for variants with no missing genotypes; the
  code falls back to the legacy pairwise scanner when missingness is present.
- Dense LD for ColocBoost/SuSiE should still use block/tiled matrix kernels in a
  future pass; this benchmark covers the current `plink_ld()` pair-table API.
